---
title: "A3 - Incarceration Trends"
author: "Kara Pugh"
date: '2022-05-22'
output: html_document
---

### Introduction

In the United States racism and racial prejudice have consistently been issues that negatively
affect minorities. One institution in particular where the affects of these issues can be clearly
seen is the United States' prison system. In the US, African Americans are viewed as "dangerous" and
more likely to commit crimes which in turn has led to African Americans being incarcerated much more often than people of other races.
For my analysis, I have chosen to specifically focus on data from US prisons since Also, I have
chosen to further narrow down my data by just looking at the prisons in Texas. I specifically chose
Texas because it is the state with the most prisons and the most prisoners. The variables that I have chosen to analyze are the number of prisoners for each race. I wanted to see how race affected the
overall population in prison. In my dataset, I kept the total populations of each race from 15 to 64 
in order to see how the number of prisoners in each race compared to the overall total of people of
that race. I also kept the total number of prisoners so that I could compare how each race compared 
to the overall prison population. By analyzing these variables I hoped to gain a better understanding
of how the prison system is affected by race and how certain patterns regarding race unveil a 
problematic correlation between race and incarceration that should not be ignored.

```{r setup, include = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(usmap)

incarceration_trends <- read.csv("https://raw.githubusercontent.com/vera-institute/incarceration-trends/master/incarceration_trends.csv")
prison_race <- incarceration_trends %>%
  filter(state == "TX") %>%
  select(year, county_name, fips, aapi_prison_pop, black_prison_pop, 
         latinx_prison_pop, native_prison_pop, white_prison_pop,
         total_prison_pop, aapi_pop_15to64, black_pop_15to64, 
         latinx_pop_15to64, native_pop_15to64, white_pop_15to64)

prison_race <- na.omit(prison_race)
prison_pop_means <- prison_race %>%
  summarise_at(vars("native_prison_pop", "aapi_prison_pop", "black_prison_pop", "latinx_prison_pop", "white_prison_pop"), mean)
aapi_mean <- prison_pop_means[1, "aapi_prison_pop"]
black_mean <- prison_pop_means[1, "black_prison_pop"]
latinx_mean <- prison_pop_means[1, "latinx_prison_pop"]
native_mean <- prison_pop_means[1, "native_prison_pop"]
white_mean <- prison_pop_means[1, "white_prison_pop"]

# What proportion of people from each race make up the the prison population?

prison_pop_prop <- prison_race %>%
  mutate(aapi_pop_prop = aapi_prison_pop/total_prison_pop) %>%
  mutate(black_pop_prop = black_prison_pop/total_prison_pop) %>%
  mutate(latinx_pop_prop = latinx_prison_pop/total_prison_pop) %>%
  mutate(native_pop_prop = native_prison_pop/total_prison_pop) %>%
  mutate(white_pop_prop = white_prison_pop/total_prison_pop) %>%
  select(year, county_name, aapi_pop_prop, black_pop_prop, latinx_pop_prop, native_pop_prop, white_pop_prop)

prison_pop_prop[is.na(prison_pop_prop)] <- 0
prison_pop_prop_means <- prison_pop_prop %>%
  summarise_at(vars("aapi_pop_prop", "black_pop_prop", "latinx_pop_prop", "native_pop_prop", "native_pop_prop", "white_pop_prop"), mean)
aapi_pop_prop_mean <- prison_pop_prop_means[1, "aapi_pop_prop"]
black_pop_prop_mean <- prison_pop_prop_means[1, "black_pop_prop"]
latinx_pop_prop_mean <- prison_pop_prop_means[1, "latinx_pop_prop"]
native_pop_prop_mean <- prison_pop_prop_means[1, "native_pop_prop"]
white_pop_prop_mean <- prison_pop_prop_means[1, "white_pop_prop"]

# What is the proportion of people in prison for each race?

prison_proportions <- data.frame(prison_race)

prison_proportions <- prison_race %>%
  mutate(aapi_prop = aapi_prison_pop/aapi_pop_15to64) %>%
  mutate(black_prop = black_prison_pop/black_pop_15to64) %>%
  mutate(latinx_prop = latinx_prison_pop/latinx_pop_15to64) %>%
  mutate(native_prop = native_prison_pop/native_pop_15to64) %>%
  mutate(white_prop = white_prison_pop/white_pop_15to64) %>%
  select(year, county_name, aapi_prop, black_prop, latinx_prop, native_prop, white_prop)

# What is the maximum proportion of people in prison for each race?

prison_proportions[is.na(prison_proportions)] <- 0
aapi_max_prop <- max(prison_proportions$aapi_prop)
black_max_prop <- max(prison_proportions$black_prop)
latinx_max_prop <- max(prison_proportions$latinx_prop)
native_max_prop <- max(prison_proportions$native_prop)
white_max_prop <- max(prison_proportions$white_prop)

# Which county has the most black prisons? The highest black proportion?

most_black_prisoners <- prison_race %>%
  filter(black_prison_pop == max(prison_race$black_prison_pop)) %>%
  pull(county_name, black_prison_pop)
highest_black_prop <- prison_proportions %>%
  filter(black_prop == black_max_prop) %>%
  pull(county_name, black_prop)

counties_prison_pop <- prison_race %>%
  select(year, county_name, fips, black_prison_pop) %>%
  group_by(county_name) %>%
  filter(year == max(year))

# Which county has the most prisoners?
most_prisoners <- prison_race %>%
  filter(total_prison_pop == max(prison_race$total_prison_pop)) %>%
  pull(county_name)

# data for just Harris county
harris_county <- prison_race %>%
  filter(county_name == "Harris County") %>%
  select(year, county_name, black_prison_pop, white_prison_pop)

# data for just Anderson county
anderson_county <- prison_proportions %>%
  filter(county_name == "Anderson County") %>%
  select(year, county_name, black_prop, white_prop)

```


### Summary Information

In the Texas prison systems the average number of black prisoners is much **higher** than the average
number of prisoners for other races with the mean for black prisoners being `r black_mean` and the
mean for latinx prisoners, white prisoners, Asian American/Pacific Islander prisoners, and native 
prisoners being `r latinx_mean`, `r white_mean`, `r aapi_mean`, and `r native_mean` respectively. 
The average proportion of black prisoners compared to the total prison population 
(black prisoners/total prisoners) is `r black_pop_prop_mean` which is significantly higher than both
Asian American/Pacific Islander prisoners (`r aapi_pop_prop_mean`) and native prisoners (`r native_pop_prop_mean`). But not quite as high as white prisoners (`r white_pop_prop_mean`) and latinx
prisoners (`r latinx_pop_prop_mean`) since white and latinx make up a vast majority of the population
in Texas. However, when we look at each race individually the data is no longer skewed by the larger general populations of whites and latinx and we see that African Americans are disproportionately incarcerated. Comparing the number of people from each race in prison versus that race's overall population per county (ie. black prisoners/black population) we see that the proportion of the black population in prison is much higher than the proportion for other races. The maximum proportion of black people in prison is `r black_max_prop` while the maximum proportion of people in prison for white and latinx people does not even reach 0.1 even though they make up the majority of Texas' population.

### Charts

I chose to include this chart for two reasons. First, it displays how across multiple races the number
of prisoners is declining. Secondly, it visually shows how the number of black prisoners is consistently higher than the number of white prisoners. Also, I specifically chose Harris County because it is the county with the most prisoners.

```{r, echo=FALSE}

harris_county %>%
  tail(14) %>%
  ggplot( aes(x=year)) +
  geom_line(aes(y=black_prison_pop), color="blue") +
  geom_line(aes(y=white_prison_pop), color="red") +
  ylab("Prison Population") +
  geom_point(aes(y=black_prison_pop), shape=21, color="black", fill="#69b3a2", size=6) +
  geom_point(aes(y=white_prison_pop), shape=24, color="black", fill="#E69F00", size=6) +
  theme(legend.position = "right") +
  ggtitle("Black and White Prisoners in Harris County")

```


- Races
- Black = teal circle
- White = yellow triangle

I chose to include this chart because it displays how the proportion of black prisoners relates to the proportion of white prisoners in Anderson county. Also, from the chart we can see that for the most part the proportion of white people does not go past a certain level, while the proportion of black people is distributed throughout the graph.

```{r, echo=FALSE}
ggplot(anderson_county, aes(x=black_prop, y=white_prop)) + 
  geom_point(
    color="orange",
    fill="#69b3a2",
    shape=21,
    alpha=0.5,
    size=6,
    stroke = 2
  ) +
  xlab("Proportion of Black People in Prison") +
  ylab("Proportion of White People in Prison")

```

### Map

I chose to include this map because it provides a good visualization of how many black prisoners there are in each Texas county. From the map it is clear that there are some counties that have a very large number of black prisoners and some that do not have a very many. In other words, the distribution is concentrated in a few specific counties. 

```{r, echo=FALSE}

plot_usmap(data = counties_prison_pop, region = "counties", values = "black_prison_pop", include = "TX", color = "blue") +
  scale_fill_continuous(low = "white", high = "blue", name = "Black Prison Population", label = scales::comma) + 
  labs(title = "Texas Black Prison Populations per County", subtitle = "using the most recent year available for each county") +
  theme(legend.position = "right")
```
